Realizing Temporal Graphs

### 2023-01-13 RealizingTemporalGraphs.tex
FPT algorithm proof first draft. Need to still add inside George's proof + introduce the definition of line segments.

In the following we prove that knowing the structure (the exact underlying path) of a fastest temporal path $P$, among two vertices of interest $u,v \in U$,
results in determining the labeling of each edge in the fastest temporal path from $u$ to $v$ (with the exception of some constant number of edges), up to the label of the first edge.

@@

After finishing the second step of the FPT algorithm, labels on all edges of fastest temporal paths from $u$ to $v$, where $u,v \in U$, are determined up to the label of the first edge.
There may still be some edges in the graph, that are unlabeled. Note that these edges are exactly the ones that are not a part of any fastest temporal path.
Let us first observe some properties of such edges.

@@

From the above it follows that there are at most $f(k)$ (namely $O(|U|^2$) unlabeled paths in $G'$ after algorithm finishes step 2.
Now we want to extend the labeling from Step 2 to all unlabeled paths.
To do this we want to relate the labeling of an unlabeled path to some fully labeled (\ie fastest temporal) path.
Let us first observe some properties.

@@

Since $x$ is an inner vertex of $P$ and is of degree two, then we know the labels of its two edges incident to it.
Denote with $\ell_x$ the edge incident to $x$ that is closer to $u'$ on the path $P$, and with $r_x$ the edge that is closer to $v'$ on $P$.
Let $y$ be an inner vertex on the path $Q$.
Similarly, denote with $\ell_y$ the edge incident to $y$ that is closer to $u$ on the path $Q$, and with $r_y$ the edge that is closer to $v$ on $Q$.
Observe the following. 
To reach $y$ from $x$ there are $4$ different options, at the start path uses either edge $\ell_y$ or $r_y$, 
and at the end path must use either $\ell_x$ or $y_x$.
Since this is true for any path (hence any temporal path), it holds also for the fastest temporal pathfrom $x$ to $y$.

@@

\begin{lemma}

@@

    Then we can precisely determine the labels on every edge of $Q$, but one, up to the label $c_{u'v'}$ of the first edge of path $P$.

@@

\subsubsection*{Step 3(a) of the FPT algorithm}

\subsection{Step 6 of the FPT algorithm}
\todo[inline]{This section is wrong, we keep this here as we will use some of the ideas.}
Let $\lambda$ be the labeling of $G'$, that satisfies $D'$.
Let $s \in V(G) \setminus V(G')$, be a vertex of degree one, that was removed from $G$ in Step 1,
and let $u$ be its neighbor in $G$.
First, we can easily assume that $|N_{G'}(u)|\geq 1$, if this is not true, then either $G$ is a star or $G'$ is disconnected and therefore also $G$ is disconnected, where $u$ is in a part of $G$ that forms a star.
We can solve the problem for stars in polynomial time (see \cref{thm:deltaExact-PolyTimeTrees}).

To determine the correct label of edge $su$ we do the following.
We find vertex $v_i \in N_{G'}(u)$ that is the closest to $v$, regarding the values of $D$, \ie  $D_{s,v_i} \leq D_{s,v_j}$ for all $v_j \in N_{G'}(u)$.
We claim, that the path $P=(s,u,v_i)$ has to be the fastest temporal path from $s$ to $v_i$ in $(G, \lambda'')$.
Suppose that is not true, then there exists a faster path $P^*$ from $s$ to $v_i$, that goes through $u$ (as this is the only neighbor of $s$), through some other vertex $v_j \in N_{G'}(u) \setminus \{v_i \}$, and through some other path $P'$ in $G$, before it finishes it $v_i$, where $P'$ is at least an edge (from $v_j$ to $v_i$). 
Therefore $P^* = (s,u,v_j,P',v_i)$, where $d(P^*) \leq d(P)$, now since $D_{s,v_i} \leq D_{s,v_j}$ for all $v_j \in N_{G'}(u)$, the first part of path $P^*$ from $s$ to $v_j$ takes at least $D_{s,v_i}$ time,
since $v_i \neq v_j$, we need at least one more time-step (one more edge) to traverse from $v_i$ to reach $v_j$, so $d(P^) \geq D_{s,v_i} + 1$, and so $P^*$ cannot be faster than $P$.
Since $P=(s,u,v_i)$ is the fastest temporal path from $s$ to $v_i$ we calculate the label of edge $su$ as $\lambda (s,v_i) \equiv \lambda(u v_i) + 1 - D_{s,v_i} \pmod \Delta$.

We repeat the above procedure for all degree one vertices in $G$, that we removed in Step 1, and we end up with the labeling $\lambda$ of the whole graph.
Now we have to make sure that all the new labels satisfy $D$, to do this we calculate the matrix $D'$ of fastest temporal paths among all pairs of vertices in $(G, \lambda)$ and compare the results with $D$, if any values differ, we stop and return a \textit{NO} answer, if $D = D'$ we return the labeling $\lambda$.

To calculate $\lambda (s,v_i)$ we need $O(|N_{G'}(u)|) = O(n)$ time, therefore to calculate all labels of edges incident to vertices of degree one in $G$, we need $O(n^2)$ time, calculating $D'$ takes $O(n^3\log n)$ time (see \cref{lemma:calculatingD}), therefore the whole step is performed in polynomial time.

>>>>

In the following we prove that knowing the structure (the exact underlying path) of a fastest temporal path $P$, among two vertices of interest $u,v \in U$,
results in determining the labeling of each edge in the fastest temporal path from $u$ to $v$ (with the exception of some constant number of edges), with respect to the label of the first edge.

@@

After finishing the second step of the FPT algorithm labels on all edges of fastest temporal paths from $u$ to $v$, where $u,v \in U$, are determined with respect to the label of the first edge.
There may still be some edges in the graph, that are unlabeled. Note that these edges are exactly the ones that are not a part of any fastest temporal path.
Let us first observe some properties of such edges.

@@

From the above it follows that there are at most $f(k)$ (namely $O(|U|^2$) unlabeled paths in $G'$ after algorithm finishes Step 2.
Now we want to extend the labeling from Step 2 to all unlabeled paths.
To do this we want to relate the labeling of an unlabeled path to some labeled (\ie fastest temporal) path.
Let us first observe some properties.

@@

Since $x$ is an inner vertex of $P$ and is of degree two, then we know the labels of its two edges incident to it.
Denote with $\ell_x$ the edge incident to $x$ that is closer to $u'$ on the path $P$, and with $r_x$ the edge that is closer to $v'$ on $P$.
Let $y$ be an inner vertex on the path $Q$.
Similarly, denote with $\ell_y$ the edge incident to $y$ that is closer to $u$ on the path $Q$, and with $r_y$ the edge that is closer to $v$ on $Q$.
Observe the following. 
To reach $y$ from $x$ there are $4$ different options, at the start path uses either edge $\ell_y$ or $r_y$, 
and at the end path must use either $\ell_x$ or $y_x$.
Since this is true for any path (hence any temporal path), it holds also for the fastest temporal path from $x$ to $y$.

@@

\begin{lemma}\label{lem:FPT-DetermineLabelsOnUnlabeledNonFastestPaths}

@@

    Then we can precisely determine the labels on every edge of $Q$, but one, with respect to the label $c_{u'v'}$ of the first edge of path $P$.

@@

Up to now we have determined the labels (the constraints for labels) of edges of $G'$. We will now extend the labeling to the initial graph $G$. 
Recall, $G'$ was created from $G$ by iteratively removing all vertices of degree $1$ from $G$.

Let $s \in V(G) \setminus V(G')$ be a vertex of degree one that was removed from $G$ at the beginning,
and let $u$ be its neighbor in $G$.
First, we can easily assume that $|N_{G'}(u)|\geq 1$. If this is not true then either $G$ is a star or $G'$ is disconnected and therefore also $G$ is disconnected, where $u$ is in a part of $G$ that forms a star.iteratively
We can solve the problem for stars in polynomial time (see \cref{thm:deltaExact-PolyTimeTrees}).

To determine the correct label of edge $su$ we use the following properties.
\begin{claim}
Let $v_i \in N_{G'}(u)$ be the closest vertex to $v$, regarding the values of $D$, \ie  $D_{s,v_i} \leq D_{s,v_j}$ for all $v_j \in N_{G'}(u)$.
Then the path $P=(s,u,v_i)$ has to be the fastest temporal path from $s$ to $v_i$ in $G$.
\end{claim}
\begin{proof}
Suppose that this is not true.
Then there exists a faster path $P^*$ from $s$ to $v_i$, that goes through $u$ (as this is the only neighbor of $s$), through some other vertex $v_j \in N_{G'}(u) \setminus \{v_i \}$, 
and through some other path $P'$ in $G$, before it finishes in $v_i$, where $P'$ is at least an edge (from $v_j$ to $v_i$). 
Therefore $P^* = (s,u,v_j,P',v_i)$, where $d(P^*) \leq d(P)$.
Now since $D_{s,v_i} \leq D_{s,v_j}$ for all $v_j \in N_{G'}(u)$ the first part of path $P^*$ from $s$ to $v_j$ takes at least $D_{s,v_i}$ time.
Since $v_i \neq v_j$ we need at least one more time-step (one more edge) to traverse from $v_i$ to reach $v_j$. 
So $d(P^) \geq D_{s,v_i} + 1$, and so $P^*$ cannot be faster than $P$.
\end{proof}

Since $P=(s,u,v_i)$ is the fastest temporal path from $s$ to $v_i$ we calculate the label of edge $su$ as $\lambda (s,v_i) \equiv \lambda(u v_i) + 1 - D_{s,v_i} \pmod \Delta$.
We repeat the above procedure for all degree one vertices in $G$, that we removed in Step 1, and we end up with the labeling $\lambda$ of the whole graph.

\begin{comment}
Now we have to make sure that all the new labels satisfy $D$, to do this we calculate the matrix $D'$ of fastest temporal paths among all pairs of vertices in $(G, \lambda)$ and compare the results with $D$, if any values differ, we stop and return a \textsc{No} answer, if $D = D'$ we return the labeling $\lambda$.

To calculate $\lambda (s,v_i)$ we need $O(|N_{G'}(u)|) = O(n)$ time, therefore to calculate all labels of edges incident to vertices of degree one in $G$, we need $O(n^2)$ time, calculating $D'$ takes $O(n^3\log n)$ time (see \cref{lemma:calculatingD}), therefore the whole step is performed in polynomial time.
\end{comment}
%
%
\subsubsection*{Step 3(a) of the FPT algorithm}
Using all of the above results we proceed in the following way.
For a partial ILP instance $I$ we 
select an inner vertex $x$ of every fastest temporal path among vertices of interest $u'$ and $v'$,
and for every unlabeled path among vertices of interest $u$ and $v$ we find such $y$ and $y_1, y_2$ from this path, that $y$ is furthest away from $x$ and $y_1, y_2$ are its neighbors.
We then enumerate $4$ possible ways that a fastest temporal path starts from $x$ and reaches $y$, 
and then $2$ additional ways, how a fastest temporal path starts from $x$ to $y_1$ (resp.~$y_2$).
For each of the possibilities we then precisely determine labels of the unlabeled path, up to the constant of the fastest $u',v'$ temporal path.

Since there are at most $|U|^2$ fastest temporal paths among vertices of interest, we create at most $4|U|^2$ extra instances for each unlabeled path.
Therefore one partial ILP instance $I$ is further extended into $O(|U|^3)=f(k)$ \todo{$O(2^k)$} final ILP instances $I_I$.

\subsubsection*{Step 3(b) of the FPT algorithm}
Using results from \cite{Lenstra1983Integer} we know that an ILP instance can be solved in FPT time with respect to the number of variables. In the case when the input instance as a solution, algorithm also provids us with one.

One final ILP instance $I_I$, that was built by our algorithm, has $f(k)$ variables. 
Therefore one $I_I$ can be solved in FPT time with respect to $k$.

We have build $f(k)$ different final ILP instances. If none of the instances is a \textsc{Yes} instance, then we conclude that there is no labeling $\lambda$ of $G$, that would satisfy the distances of the matrix $D$,
and therefore $D$ is a \textsc{No} instance of \deltaExact.
If there is a final ILP instance $I_I$, that is a \textsc{Yes} instance, then we take a solution $S$, provided by the algorithm from \cite{Lenstra1983Integer}, and determine the corresponding labeling $\lambda$ of $G$.
Note, the minimum number of solutions of $I_I$ is at least $\delta$, since the graph is periodic and adding the same constant to every edge in the graph, dose not change the duration of any temporal path in $(G, \lambda)$.
